---
title: "Build a Citation Scanner"
vignette: >
  %\VignetteIndexEntry{Build a Citation Scanner}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
library(ascribe)
```

For a complete, real-world implementation of a package citation scanner built with `ascribe`, see [stanflow](https://github.com/VisruthSK/stanflow).

`ascribe` separates AST parsing from citation generation. This guide demonstrates building a citation scanner targeting base R packages (`stats`, `graphics`, `utils`, `methods`).

## Build the package universe

`build_universe_data()` extracts exported functions, R6 methods, and origin mappings for re-exported functions.

```{r}
base_pkgs <- c("stats", "graphics", "utils", "methods")
universe <- build_universe_data(base_pkgs)
names(universe)
```

Run `build_universe_data()` in an environment where all target packages are installed.

## Scan source code

`scan_usage()` inspects `.R`, `.Rmd`, and `.qmd` files for package attachments (`library()`, `require()`) and function calls (`pkg::fun()` or unqualified calls resolved by attachment order).

```{r}
project <- tempfile("base-project-")
dir.create(project)
script <- file.path(project, "analysis.R")

writeLines(
  c(
    "library(stats)",
    "library(graphics)",
    "fit <- lm(mpg ~ wt, data = mtcars)",
    "summary(fit)",
    "plot(fit)",
    "utils::capture.output(summary(fit))"
  ),
  script
)

usage <- scan_usage(
  project,
  universe,
  strict = TRUE
)

usage
unlink(project, recursive = TRUE)
```

`strict = TRUE` aborts on calls that cannot be resolved unambiguously; the default, `strict = FALSE`, warns and omits them instead.

## Generate citations

`cite_usage()` formats BibTeX entries for detected packages and functions.

```{r}
citations <- cite_usage(usage)
writeLines(citations)
```

Override default package or function citations with `package_citations` and `function_citations`.

## Save scanner data in sysdata.rda

`generate_universe_sysdata()` precompiles universe data into `R/sysdata.rda` for downstream package distribution.

```{r}
#| eval: false
# data-raw/sysdata.R
generate_universe_sysdata(
  packages = c("stats", "graphics", "utils", "methods"),
  prefix = "base",
  file = "R/sysdata.rda"
)
```

This saves `.base_pkgs`, `.base_exports`, `.base_export_index`, `.base_origin_map`, and `.base_pkg_versions` for fast loading without runtime package inspection.